# Week 3 Assessment

First, load the tidyverse and import data.

```{r}

library(tidyverse)

bor_nestlings <- read_csv(here::here("data/73_species.csv"))
bor_traits <-read_csv(here::here("data/Traits_73_species.csv"))
avonet <- readxl::read_excel(here::here("data/AVONET Supplementary dataset 1.xlsx", sheet = "AVONET1_BirdLife"))
```

Now, lets explore the data together: *Are nestlings showing up earlier in the year over time?*

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Sure looks like they are showing up earlier each year.

Let's see the breakdown by species.

```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

So chaotic! What if we just look at the 5 most data-rich species?

```{r}
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

A general pattern, with one conflicting example. What species have the strongest trends?

```{r}

# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))
```

Here are 5 of the species with the strongest trends:

```{r}
strongest <- bor_trends %>% 
  arrange(-abs(doy_trend))

strongest[1:5,]
```

Lets plot of a few of these species:

```{r}
soi <- c("ARDCIN", "LARMIN")
species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Find the species with the most positive and most negative trends, then plot just those.

```{r}
nrow_bor_trends <- nrow(bor_trends) # Use this later

bor_extreme <- bor_trends %>% 
  # Sort by the day of year trend
  arrange(doy_trend) %>% 
  # Keep just the first (most negative trend) and last (most positive trend) rows
  slice(c(1, nrow_bor_trends))

# Now plot them
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

Next we\'re going to test the hypothesis that smaller birds have more flexible phenology, i.e. the absolute value of the trend is correlated with smaller body sizes.

To test our hypothesis, we need to add AVONET data to our phenology data by joining. The keys are a bit tricky here!

```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = `Scientific name`)
```
