# Week 3 Assessment

First, load the tidyverse and import data.

```{r}
library(tidyverse)

bor_nestlings <- read_csv(here::here("data/73_species.csv"))
bor_traits <-read_csv(here::here("data/Traits_73_species.csv"))
avonet <- readxl::read_excel(here::here("data/AVONET Supplementary dataset 1.xlsx"))
```

Now, lets explore the data together: *Are nestlings showing up earlier in the year over time?*

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Sure looks like they are showing up earlier each year.

Let\'s see the breakdown by species.

```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

So chaotic! What if we just look at the 5 most data-rich species?

```{r}
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

A general pattern, with one conflicting example. What species have the strongest trends?
